The Radio Sky at Meter Wavelengths: $m$-Mode Analysis Imaging with the Owens Valley Long Wavelength Array

### 2017-06-27 paper.tex
reorder sections
mostly elaborating on correlated rfi removal

\author{
    Michael W. Eastwood,
    Gregg Hallinan
}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figures/antenna-layout}
    \caption{
        This figure shows the antenna layout for the OVRO-LWA. Black dots correspond to antennas
        within the 200 m diameter core of the array. The 32 open circles are the expansion antennas
        built in early 2016 in order to increase the longest baseline to 1.5 km. The 5 crosses are
        antennas equipped with noise-switched front ends.
    }
    \label{fig:antenna-layout}
\end{figure}

The flux scale is tied to the work of \citet{2016arXiv160905940P}.

Without removing bright sources from the data, sidelobes from bright sources will dominate the
variance in the image.  At 74 MHz Cyg A is a 15,000 Jy source \citep{2016arXiv160905940P}. A
conservative estimate for the confusion limit at 74 MHz with a 10 arcminute beam is 500 mJy
\citep{2012RaSc...47.0K04L}. Therefore we require that Cyg A's sidelobes be at most of $-$45 dB down
from its peak flux to prevent Cyg A's sidelobes from dominating the variance in the image.
\todo{Measure Cyg A's sidelobe levels if it is not subtracted}

\subsubsection{Sparking Power Lines}

The Owens Valley is an important source of water and power for the city of Los Angeles.
Unfortunately this means that high voltage power lines run along the valley to the west of the OVRO-
LWA. Some of these power line poles have faulty insulators that arc and produce pulsed, broadband
RFI. Because these poles exist in the near-field of the array, we have been able to localize some of
them by using the curvature of the incoming wavefront. Efforts are currently underway to work with
the utility pole owners to have these insulators replaced.

In the meantime it is possible to suppress their contamination in the dataset. The contribution of
these RFI sources to the visibilities can be plainly seen by averaging $>$24 hours of data with the
phase center set to zenith. In this way, true sky components are smeared along tracks of constant
declination while terrestrial sources (ie. the arcing power lines) are not smeared. For similar
reasons, when constructing a map of the sky from $>$24 hours of data these sparking power lines will
produce artifacts in the map that are smeared along tracks of constant declination. However while
the amplitude of the RFI will fluctuate with time, its phase will be constant. Therefore if you can
use isolate a set of model visibilities for the RFI, it can be removed from each integration by
scaling and subtracting these model visibilities from the measured visibilities.

Obtaining a model for the RFI is complicated by the fact that the contaminating sources are at
extremely low elevations where the antenna response is maximally inhomogeneous. Therefore it is not
enough to know the physical location of the faulty insulator generating the RFI. In addition you
must also know the response of each antenna (amplitude and phase) in the appropriate direction. This
motivates the use of peeling, which allows the antenna response to be a free parameter.

Therefore model visibilities for the RFI can be obtained by peeling the sources after smearing the
visibilities over $>$24 hours. These model visibilities are then scaled and subtracted from each
integration in the data set.

\todo{show an example smeared image before and after peeling}

\section{Imaging}

@@

Equation~\ref{eq:m-mode-sum-equation} can be recognized as a matrix equation where the transfer
matrix $\b B$ is block-diagonal.
\begin{equation}
    \b B = \left(\begin{array}{cccc}
        m = 0 &&& \\
              & m=\pm1 && \\
              && m=\pm2 & \\
              &&& \ddots \\
    \end{array}\right)
\end{equation}
The vector $\b v$ contains the list of $m$-modes and the vector $\b a$ contains the list of
spherical harmonic coefficients representing the sky brightness. In order to take advantage of the
block-diagonal structure in $\b B$, $\b v$ and $\b a$ must be sorted by the value of $m$.
\begin{equation}\label{eq:m-mode-matrix-equation}
    \overbrace{\left(
        \begin{array}{c}
            \vdots \\
            m\text{-modes} \\
            \vdots \\
        \end{array}
    \right)}^{\b v}
    =
    \overbrace{\left(
        \begin{array}{ccc}
            \ddots & & \\
            & \text{transfer matrix} & \\
            & & \ddots \\
        \end{array}
    \right)}^{\b B}
    \overbrace{\left(
        \begin{array}{c}
            \vdots \\
            a_{lm} \\
            \vdots \\
        \end{array}
    \right)}^{\b a}
\end{equation}

@@

In practice, the L-curve should be used as a guide to estimate a reasonable value of $\epsilon$.
However better results can often be obtained by tuning the value of $\epsilon$. For instance
increasing the value of $\epsilon$ can improve the noise properties of the map by down-weighting
noisy modes. Decreasing the value of $\epsilon$ can improve the resolution in the map by
up-weighting the contribution of longer baselines, which are likely fewer in number. In this respect
choosing the value of $\epsilon$ is analagous to picking the weighting scheme in traditional
imaging where robust weighting schemes can be tuned to similar effect. \todo{citation}

@@

\subsection{Image Deconvolution}

@@

\section{Errors}

\subsection{Terrestrial Interference}

When writing down equation~\ref{eq:basic-imaging}, it is implicitly assumed that the correlated
voltage fluctuations measured between pairs of antennas are generated by astronomical sources of
radio emission. In practice, this implicit assumption can be violated. For instance a low-frequency
interferometer located in the vicinity of an arcing power line will see an additional contribution
from the radio-frequency interference (RFI) generated by the arcing process. Similarly common-mode
pickup along the analog signal path of the interferometer may generate an additional spurious
contribution to the measured visibilities. While the amplitude and phase of these contaminating
signals may fluctuate with time, they do not sweep across the sky at the sidereal rate
characteristic of astronomical sources.

The first step in equation~\ref{eq:tikhonov-solution} is to compute $\b B^*\b v$. In this step we
compute the projection of the measurement $\b v$ onto the space spanned by the columns of $\b B$.
Each column of $\b B$ describes the interferometer's response to a corresponding spherical harmonic
coefficient of the sky brightness distribution. Therefore the act of computing $\b B^*\b v$ is to
project the measured $m$-modes onto the space of $m$-modes which could be generated by astronomical
sources. The degree to which a source of terrestrial interferer will contaminate a map generated
using $m$-mode analysis imaging is determined by its amplitude after projection.

For instance, a bright interfering source might contribute $\b v_\text{terrestrial}$ to the measured
$m$-modes. However, if $\b v_\text{terrestrial}$ is actually perpendicular to all of the columns of
$\b B$, there will be no contamination in the map because $\b B^*\b v_\text{terrestrial} = 0$.  In
practice this is unlikely. In general the contamination is proportional to the overall amplitude of
the interference ($\|\b v_\text{terrestrial}\|$) and the degree to which the interference mimics
an astronomical signal ($\|\b B^*\b v_\text{terrestrial}\|/\|\b v_\text{terrestrial}\|$).

These astronomical sources are anchored to the celestial sphere and therefore rotate
across the sky over the course of a sidereal day. Terrestrial source of interference are not
anchored to the celestial sphere.

This characteristic motion of astronomical sources
is a fundamental component to $m$-mode analysis imaging

radio emission is astronomical. Astronomical sources are anchored to the celestial sphere and
therefore rotate across the sky over the course of a sidereal day. However, in a practice

$m$-mode analysis imaging implicitly assumes that the correlation between any two pairs of antennas

However it is instructive to analyze the contaminating effect the RFI will have
on $m$-mode analysis maps of the sky. The correlation observed on a given
baseline and frequency channel for a single source of RFI is
\begin{equation}
    V(t) = I(t) B^\textrm{near-field}
\end{equation}

However we need a strategy to mitigate the RFI before imaging. A constant source attached to the
horizon has a fringe rate of zero. The only possible astronomical origins for a source with zero
fringe rate are:

\begin{enumerate}
    \item a point source located at the north or south celestial poles, and
    \item a ring of constant declination.
\end{enumerate}

Therefore while snapshot images will show the RFI source on the horizon, extended integrations will
show the RFI as rings of constant declination. We can take advantage of this discrepancy to separate
astronomical emission from terrestrial RFI.

@@

\subsection{Ionospheric Scintillation}

One of the key assumptions made by $m$-mode analysis is that the sky is completely static.  We
assume that the only time-dependent behavior is the rotation of the Earth, which slowly rotates the
sky through the fringe patterns of the interferometer. At low frequencies the ionosphere violates
this assumption. In particular, ionospheric scintillation will cause even static sources to exhibit
significant variability.

The correlation observed on a given baseline for a single point source is
\begin{equation}
    V_\nu(t_{\textrm{sidereal}}) = I_\nu B_\nu(t_{\textrm{sidereal}}),
\end{equation}
where $I_\nu$ is the flux of the source at the frequency $\nu$, and $B_\nu$
is the baseline transfer function. The transfer function is a function of
the direction to the source, which is in turn a function of the sidereal
time $t_{\textrm{sidereal}}$. If the source is varying, from intrinsic
variability or due to scintillation, than the source flux is also a function
of the time coordinate $t$ such that
\begin{equation}
    V_\nu(t_{\textrm{sidereal}}) = I_\nu(t) B_\nu(t_{\textrm{sidereal}}),
\end{equation}
where $t_{\textrm{sidereal}} = (t \mod 23.9345\,\textrm{hours})$.

Now assume we have observed with our interferometer for a single sidereal day.
In order to compute the $m$-modes we must Fourier transform with respect to
sidereal time. In a real measurement this is a discrete Fourier transform of
the observed correlation with respect to time (where the sum over time is
restricted to a sidereal day). \todo{fix normalization}
\begin{equation}
    V_{\nu, m} = \sum_{t} V_\nu(t) e^{-imt}
\end{equation}
Define $V_{\nu, m}^{\textrm{static}}$ to be the observed $m$-modes if the source
was actually static ($I_\nu(t) \equiv I_{\nu,0}$). Then as a consequence of the
Fourier convolution theorem
\begin{equation}
    V_{\nu, m} = \sum_{m^\prime} V_{m^\prime}^\textrm{static} I_{\nu, m-m^\prime}.
\end{equation}

This will tend to scatter power between $m$-modes.

Simulate images of point sources with this

\subsection{Ionospheric Refraction}

@@

\begin{figure*}[ht]
    \includegraphics[width=\textwidth]{maps/rgb-map-small}
    \caption{three color map}
\end{figure*}

\begin{figure*}[ht]
    \centering
    \begin{tabular}{cc}
        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} &
        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} \\
        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} &
        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} \\
        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} &
        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} \\
        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} &
        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} \\
    \end{tabular}
    \caption{single channel maps}
\end{figure*}

\begin{figure*}[ht]
    \centering
    \begin{tabular}{cc}
        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} &
        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} \\
        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} &
        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} \\
        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} &
        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} \\
        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} &
        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} \\
    \end{tabular}
    \caption{jackknifed error maps}
\end{figure*}

@@

@@

These maps are pretty cool.

>>>>

\author{
    Michael W. Eastwood,
    Gregg Hallinan,
    Twilight Sparkle,
    Rainbow Dash
}

@@

\section{All-Sky Imaging}

@@

Equation~\ref{eq:m-mode-sum-equation} can be recognized as a matrix equation where the transfer
matrix $\b B$ is block-diagonal.
\begin{equation}
    \b B = \left(\begin{array}{cccc}
        m = 0 &&& \\
              & m=\pm1 && \\
              && m=\pm2 & \\
              &&& \ddots \\
    \end{array}\right)
\end{equation}
The vector $\b v$ contains the list of $m$-modes and the vector $\b a$ contains the list of
spherical harmonic coefficients representing the sky brightness. In order to take advantage of the
block-diagonal structure in $\b B$, $\b v$ and $\b a$ must be sorted by the value of $m$. Positive
and negative values of $m$ are grouped together because the brightness of the sky is real-valued,
and the spherical harmonic transform of a real-valued function has $a_{l(-m)} = (-1)^m a_{lm}^*$.
\begin{equation}\label{eq:m-mode-matrix-equation}
    \overbrace{\left(
        \begin{array}{c}
            \vdots \\
            m\text{-modes} \\
            \vdots \\
        \end{array}
    \right)}^{\b v}
    =
    \overbrace{\left(
        \begin{array}{ccc}
            \ddots & & \\
            & \text{transfer matrix} & \\
            & & \ddots \\
        \end{array}
    \right)}^{\b B}
    \overbrace{\left(
        \begin{array}{c}
            \vdots \\
            a_{lm} \\
            \vdots \\
        \end{array}
    \right)}^{\b a}
\end{equation}

@@

In practice, the L-curve should be used as a guide to estimate a reasonable value of $\epsilon$.
However better results can often be obtained by tuning the value of $\epsilon$. For instance
increasing the value of $\epsilon$ can improve the noise properties of the map by down-weighting
noisy modes. Decreasing the value of $\epsilon$ can improve the resolution in the map by
up-weighting the contribution of longer baselines, which are likely fewer in number. In this respect
choosing the value of $\epsilon$ is analagous to picking the weighting scheme in traditional
imaging where robust weighting schemes can be tuned to similar effect \citep{briggs}.

@@

\subsection{$m$-Mode Analysis Image Deconvolution}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figures/antenna-layout/antenna-layout}
    \caption{
        This figure shows the antenna layout for the OVRO-LWA. Black dots correspond to antennas
        within the 200 m diameter core of the array. The 32 open circles are the expansion antennas
        built in early 2016 in order to increase the longest baseline to 1.5 km. The 5 crosses are
        antennas equipped with noise-switched front ends.
    }
    \label{fig:antenna-layout}
\end{figure}

The flux scale is tied to the work of \citet{1977A&A....61...99B, 2012MNRAS.423L..30S,
2017ApJS..230....7P}.

Without removing bright sources from the data, sidelobes from bright sources will dominate the
variance in the image.  At 74 MHz Cyg A is a 15,000 Jy source \citep{2017ApJS..230....7P}. A
conservative estimate for the confusion limit at 74 MHz with a 10 arcminute beam is 500 mJy
\citep{2012RaSc...47.0K04L}. Therefore we require that Cyg A's sidelobes be at most of $-$45 dB down
from its peak flux to prevent Cyg A's sidelobes from dominating the variance in the image.
\todo{Measure Cyg A's sidelobe levels if it is not subtracted}

\section{Error Analysis}

\subsection{Terrestrial Interference and Pickup}

When writing down equation~\ref{eq:basic-imaging}, it is implicitly assumed that the correlated
voltage fluctuations measured between pairs of antennas are exclusively generated by astronomical
sources of radio emission. In practice this implicit assumption can be violated. For instance a
low-frequency interferometer located in the vicinity of an arcing power line will see an additional
contribution from the radio-frequency interference (RFI) generated by the arcing process. Similarly
common-mode pickup along the analog signal path of the interferometer may generate an additional
spurious contribution to the measured visibilities. While the amplitude and phase of these
contaminating signals may fluctuate with time, they do not sweep across the sky at the sidereal rate
characteristic of astronomical sources.

The Owens Valley is an important source of water and power for the city of Los Angeles.
Unfortunately this means that high voltage power lines run along the valley to the west of the OVRO-
LWA. Some of these power line poles have faulty insulators that arc and produce pulsed, broadband
RFI. Because these poles exist in the near-field of the array, we have been able to localize some of
them by using the curvature of the incoming wavefront to infer a distance. Efforts are currently
underway to work with the utility pole owners to have these insulators replaced.

In the meantime it is possible to suppress their contamination in the dataset. The contribution of
these RFI sources to the visibilities can be plainly seen by averaging $>24$ hours of data with the
phase center set to zenith. In this way, true sky components are smeared along tracks of constant
declination while terrestrial sources (ie. the arcing power lines or any common-mode pickup) are not
smeared.  Obtaining a model for the RFI is complicated by the fact that the contaminating sources
are at extremely low elevations where the antenna response is essentially unknown (and inhomogeneous
due to antenna shadowing effects). Therefore it is not enough to know the physical location of the
faulty insulator generating the RFI. In addition you must also know the response of each antenna
(amplitude and phase) in the appropriate direction. This motivates the use of peeling
\citep{2008ISTSP...2..707M, 2015MNRAS.449.2668S}, which allows the antenna response to be a free
parameter.  Therefore model visibilities for the RFI can be obtained by peeling the sources after
smearing the visibilities over $>24$ hours.

%These model visibilities are then scaled and subtracted from each
%integration in the data set.
%
%
%
%For similar
%reasons, when constructing a map of the sky from $>24$ hours of data these sparking power lines will
%produce artifacts in the map that are smeared along tracks of constant declination. However while
%the amplitude of the RFI will fluctuate with time, its phase will be constant. Therefore if you can
%use isolate a set of model visibilities for the RFI, it can be removed from each integration by
%scaling and subtracting these model visibilities from the measured visibilities.
%
%
%
%
%
%
%The first step in equation~\ref{eq:tikhonov-solution} is to compute $\b B^*\b v$. In this step we
%compute the projection of the measurement $\b v$ onto the space spanned by the columns of $\b B$.
%Each column of $\b B$ describes the interferometer's response to a corresponding spherical harmonic
%coefficient of the sky brightness distribution. Therefore the act of computing $\b B^*\b v$ is to
%project the measured $m$-modes onto the space of $m$-modes which could be generated by astronomical
%sources. The degree to which a source of terrestrial interferer will contaminate a map generated
%using $m$-mode analysis imaging is determined by its amplitude after projection.
%
%For instance, a bright interfering source might contribute $\b v_\text{terrestrial}$ to the measured
%$m$-modes. However, if $\b v_\text{terrestrial}$ is actually perpendicular to all of the columns of
%$\b B$, there will be no contamination in the map because $\b B^*\b v_\text{terrestrial} = 0$.  In
%practice this is unlikely. In general the contamination is proportional to the overall amplitude of
%the interference ($\|\b v_\text{terrestrial}\|$) and the degree to which the interference mimics
%an astronomical signal ($\|\b B^*\b v_\text{terrestrial}\|/\|\b v_\text{terrestrial}\|$).
%
%
%
%
%
%
%
%
%
%
%
%
%
%
%
%
%
%These astronomical sources are anchored to the celestial sphere and therefore rotate
%across the sky over the course of a sidereal day. Terrestrial source of interference are not
%anchored to the celestial sphere.
%
%This characteristic motion of astronomical sources
%is a fundamental component to $m$-mode analysis imaging
%
%radio emission is astronomical. Astronomical sources are anchored to the celestial sphere and
%therefore rotate across the sky over the course of a sidereal day. However, in a practice 
%
%$m$-mode analysis imaging implicitly assumes that the correlation between any two pairs of antennas
%
%
%However it is instructive to analyze the contaminating effect the RFI will have
%on $m$-mode analysis maps of the sky. The correlation observed on a given
%baseline and frequency channel for a single source of RFI is
%\begin{equation}
%    V(t) = I(t) B^\textrm{near-field}
%\end{equation}
%
%However we need a strategy to mitigate the RFI before imaging. A constant source attached to the
%horizon has a fringe rate of zero. The only possible astronomical origins for a source with zero
%fringe rate are:
%
%\begin{enumerate}
%    \item a point source located at the north or south celestial poles, and
%    \item a ring of constant declination.
%\end{enumerate}
%
%Therefore while snapshot images will show the RFI source on the horizon, extended integrations will
%show the RFI as rings of constant declination. We can take advantage of this discrepancy to separate
%astronomical emission from terrestrial RFI.
%
%
%
%
%
%
%
%\todo{show an example smeared image before and after peeling}

@@

\subsection{The Ionosphere}

%One of the key assumptions made by $m$-mode analysis is that the sky is completely static.  We
%assume that the only time-dependent behavior is the rotation of the Earth, which slowly rotates the
%sky through the fringe patterns of the interferometer. At low frequencies the ionosphere violates
%this assumption. In particular, ionospheric scintillation will cause even static sources to exhibit
%significant variability.
%
%The correlation observed on a given baseline for a single point source is
%\begin{equation}
%    V_\nu(t_{\textrm{sidereal}}) = I_\nu B_\nu(t_{\textrm{sidereal}}),
%\end{equation}
%where $I_\nu$ is the flux of the source at the frequency $\nu$, and $B_\nu$
%is the baseline transfer function. The transfer function is a function of
%the direction to the source, which is in turn a function of the sidereal
%time $t_{\textrm{sidereal}}$. If the source is varying, from intrinsic
%variability or due to scintillation, than the source flux is also a function
%of the time coordinate $t$ such that
%\begin{equation}
%    V_\nu(t_{\textrm{sidereal}}) = I_\nu(t) B_\nu(t_{\textrm{sidereal}}),
%\end{equation}
%where $t_{\textrm{sidereal}} = (t \mod 23.9345\,\textrm{hours})$.
%
%Now assume we have observed with our interferometer for a single sidereal day.
%In order to compute the $m$-modes we must Fourier transform with respect to
%sidereal time. In a real measurement this is a discrete Fourier transform of
%the observed correlation with respect to time (where the sum over time is
%restricted to a sidereal day). \todo{fix normalization}
%\begin{equation}
%    V_{\nu, m} = \sum_{t} V_\nu(t) e^{-imt}
%\end{equation}
%Define $V_{\nu, m}^{\textrm{static}}$ to be the observed $m$-modes if the source
%was actually static ($I_\nu(t) \equiv I_{\nu,0}$). Then as a consequence of the
%Fourier convolution theorem
%\begin{equation}
%    V_{\nu, m} = \sum_{m^\prime} V_{m^\prime}^\textrm{static} I_{\nu, m-m^\prime}.
%\end{equation}
%
%This will tend to scatter power between $m$-modes.
%
%Simulate images of point sources with this

\subsection{Thermal Noise}

@@

%\begin{figure*}[ht]
%    \includegraphics[width=\textwidth]{maps/rgb-map-small}
%    \caption{three color map}
%\end{figure*}
%
%\begin{figure*}[ht]
%    \centering
%    \begin{tabular}{cc}
%        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} &
%        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} \\
%        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} &
%        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} \\
%        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} &
%        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} \\
%        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} &
%        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} \\
%    \end{tabular}
%    \caption{single channel maps}
%\end{figure*}
%
%\begin{figure*}[ht]
%    \centering
%    \begin{tabular}{cc}
%        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} &
%        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} \\
%        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} &
%        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} \\
%        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} &
%        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} \\
%        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} &
%        \includegraphics[width=\columnwidth, draft]{maps/rgb-map} \\
%    \end{tabular}
%    \caption{jackknifed error maps}
%\end{figure*}

@@

\section{A Nearby Supernova Remnant????}

These maps are pretty cool. Any questions?

@@
